---
title: "EX4"
author: "yizhen"
date: "2/28/2020"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.width=8, fig.height=4) 
```

##a)
```{r}
nausea = read.table("nauseatable.txt", header = TRUE)
nausea
```

```{r}
#We create a datafeam contains 2 columns: nausea, medicine
#Set patient suffered from nausea as 1, otherwise 0
nausea=c(rep(0,times=100), rep(1,times=52),rep(0, times=32), rep(1, times=35),rep(0, times=48), rep(1, times=37))

medicine=c(rep("Chlorpromazine", times=152),rep("Pentobarbital(100mg)", times=67),rep("Pentobarbital(150mg)", times=85))
newNausea = data.frame(nausea,medicine); newNausea
```

```{r}
xtabs(~medicine+nausea, data = newNausea) 
```
We could see that the result is the same as the dataset "nauseatable.txt", so the dataframe we created is correct. 

#b)
```{r}
#Permutation test
label_medic = factor(newNausea$medicine)
nausdata = data.frame(nausea=as.vector(newNausea$nausea), label_medic)

attach(nausdata)
mystat=function(x) sum(residuals(x)^2)
B=1000
tstar=numeric(B)
for(i in 1:B){
  treatstar = sample(label_medic)
  tstar[i] = mystat(lm(nausea~treatstar))
}
myt = mystat(lm(newNausea$nausea~newNausea$medicine))
hist(tstar)
myt
pl=sum(tstar<myt)/B
pr=sum(tstar>myt)/B
pl
```
We could see that p-value is 0.041, less than 0.05. Therefore, H0 should be rejected which means the medicine doesn't work equally well against nausea. 

```{r}
#chisquare test
chisq.test(xtabs(~newNausea$medicine+newNausea$nausea))[[1]]
```
By using chisquared test we could get the result which is 6.62 and we could use this to calculate the p-value. While in permutation test the result is 71.82, they use different values to calculate the final p-value.

#c)
```{r}
chisq.test(xtabs(~newNausea$medicine+newNausea$nausea))
```
We could see from chisquared test that p-value is 0.03643, smaller than 0.05, therefore, H0 should be rejected. And the p-value of both tests are close to each other. In conclusion, they use different methods but the results are the same. 
