---
title: "EX3"
author: "Yizhen"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.width=8, fig.height=4) 
```
#Exercise 3. (cow)
##a) repeated measures - exchangebale 
First we evaluate normality to determine which test to use. From the QQ-plots below we conclude that both treatment samples are normally distributed.

```{r}
cow = read.table("cow.txt", header = TRUE); 

#There are two factors influence the milk production

mm = subset(cow, treatment=="A")["milk"]
sf = subset(cow, treatment=="B")["milk"]
mm1 = as.numeric(unlist(mm))
sf1 = as.numeric(unlist(sf))

par(mfrow=c(1,2))
qqnorm(mm1)
qqnorm(sf1)
```

From the ANOVA results below, we can conclude that within-cow variation(see variable "id") the milk production differs. Because the p-value for id is less than 0.05, therefore, H0 is rejected. Furthermore, from the summary we can conclude that most of the cows(except id4) are different from the cow with id1.  
Then, we could see p-value for per is less than 0.05 so we reject H0 which means whether a cow is going through the first period or second seems to make a difference.
Furthurmore, because the p-value of treatment is 0.51654, so we do not reject H0 which means the treatment A does not significant differ from treatment B. Also this could be seen from second table (treatmentB). Therefore there is no significant difference in milk production. However, it is important to note that this is not the appropiate way of testing the cross-over design.


```{r}
cow$id = factor(cow$id)
cow$per = factor(cow$per)
cowanova = lm(milk~id+per+treatment,data = cow)
anova(cowanova); summary(cowanova)
``` 

##b) modelling the cow effect as a random effect (use the function lmer). 
TODO: add order in lmer formula
```{r}
library(lme4)
cowlmer = lmer(milk~treatment+per+(1|id), data=cow, REML = FALSE)
summary(cowlmer)
```

```{r}
cowlmerTreatment = lmer(milk~per+(1|id), data=cow, REML = FALSE)
anova(cowlmerTreatment, cowlmer)
```
Based on the p-value, we could know that treatment is not important. The result is the same as that in a). 

```{r}
cowlmerPer = lmer(milk~treatment+(1|id), data=cow, REML = FALSE)
anova(cowlmerPer, cowlmer)
```
Based on the p-value, we could know that whether a cow is going through the first treatment or second is important. 

##c)
```{r}
attach(cow)
t.test(milk[treatment=="A"],milk[treatment=="B"],paired=TRUE)
```
From this result that p-value is 0.8281 which means the treatment is not important. From previous analysis the same conclusion was reached. Given the design, it is inappropriate to use the paired t-test. Because all other factors are being ignored.