---
title: "**EDDA Group 29 Assignment 2**"
author: "Geoffrey van Driessel (12965065), Yizhen Zhao (2658811) & Sophie Vos (2551583)"
date: "3/9/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.width=8, fig.height=4) 
```

An overview of the R code is shown in the Appendix on page X.

## Exercise 1

**a)** A randomized design with two categorical factors, with (1) the first factor having three categorical levels, (2) the second factor having two levels and (3) having three samples for each unique category, can be produced with the following R code:

``` {r echo=TRUE, eval=FALSE}
I=3; J=2; N=3
rbind(rep(1:I,each=N*J),rep(1:J,N*I),sample(1:(N*I*J)))
```

**b)** The boxplot and interaction plot below confirms our intuition: (1) a cold environment causes a much slower decay, (2) wet bread has a much wider distribution (variance), (3) on average dry bread decays slower than wet bread, however, (4) wet and cold (frozen) bread has the slowest decay. From the non-parallel lines in the interaction plot and the wide distribution of the wet sample, we conclude that the (wet) humidity amplifies the effect of the temperature and it can thus be explained by the strong interaction between the two factors (opposed to the errors in the measurement).

```{r fig.width=8, fig.height=3}
bread = read.table("bread.txt", header = TRUE)
par(mfrow=c(1,3))
boxplot(hours ~ humidity, data = bread, ylab = "Hours")
boxplot(hours ~ environment, data = bread, ylab = "Hours")
interaction.plot(bread$environment, bread$humidity, bread$hours, xlab = "Temperature", ylab = "Hours", trace.label = "Humidity")
```

**c)** We have three null hyphotheses: (1) $H_0$ : there is no main effect of first factor (humidity), (2) $H_0$ : there is no main effect of second factor (environment) and (3) $H_0$ : there is no interactions between the two factors.
From the two-way ANOVA result below, we reject all null hyphotheses. This means that both factors have a main effect on the decay time of bread, and the factors have an interaction effect.

```{r}
breadaov=lm(hours~humidity*environment, data = bread); 
anova(breadaov)
```

**d)** <!-- Need review --> According to the means of squares, on average the environment has the largest effect on the decay. However, this can not easily be concluded as it is being compared to one base (the first category), instead of a more comprehensive analysis.

```{r}
summary(breadaov)$coefficients
```

**e)** The first requirement is that for each unique category, there should be at least 2 samples, which is the case. Then, the most important requirement is that the data among the factors should approximately have equal variances. This has been tested in b) and the conclusion was that they approximately were the same. A different test we can do after the ANOVA test, is check whether the error is normally distributed, which can be expected from a random variable. In the QQ-plot, it can be seen that the residuals are approximately normally distributed. In the fitted residuals plot, it can be seen that the spread is approximately horizontally symmetric among the fitted values, however, there are 2 outliers in the middle. 

```{r fig.width=8, fig.height=3}
par(mfrow=c(1,2))
qqnorm(residuals(breadaov))
plot(fitted(breadaov),residuals(breadaov))
```

## Exercise 2

**a)** The following code generates a random block design with five blocks, a factor with three levels and one sample per unique category.

```{r echo=TRUE, eval=FALSE}
B=5;
if1 = sample(1:5)
if2 = sample(6:10)
if3 = sample(11:15)
for (i in 1:B) print(c(if1[i], if2[i], if3[i]))
```

**b)** The boxplots below suggest that indeed the skill level and the interfaces matter for the search time. We see that skill level 1 and interface 1 are the fastest. From the interaction plots below, we observe clear interaction effects. Overall, the factors have the same pattern, namely, all lines start in the lower left corner and end towards the upper right corner. However, they are not perfectly parallel, this can be explained by the small sample sizes which cause local irregularities. Thus, we conclude that there is no interaction between the two factors.

```{r fig.width=8, fig.height=6}
search = read.table("search.txt", header = TRUE)
par(mfrow=c(2,2))
boxplot(time~skill, data=search)
boxplot(time~interface, data=search)
interaction.plot(search$skill,search$interface, search$time, xlab = "Skill level", ylab = "Time", trace.label = "Interface")
interaction.plot(search$interface,search$skill, search$time, xlab = "Interface", ylab = "Time", trace.label = "Skill level")
```

**c)** $H_0$ : search time is the same for all interfaces. From the ANOVA results below, it can be concluded that $H_0$ is rejected. This means that the search time is not the same for all interfaces. Furthermore, we can estimate the time it takes for a user with skill level 3 to find a product using interface 2 by looking at the summary table and adding the coefficients of these two categories to the intercept. In this case, that would be $15.015+3.033+2.7=20.748$.

```{r}
search$skill = factor(search$skill)
search$interface = factor(search$interface)
aovsearch=lm(time~interface+skill, data=search)
anova(aovsearch)
summary(aovsearch)
```

**d)** The QQ-plot of the residuals below looks normally distributed, which is good. The fitted residuals do not depict any outliers.

```{r fig.width=8, fig.height=3}
par(mfrow=c(1,2))
qqnorm(residuals(aovsearch))
plot(fitted(aovsearch),residuals(aovsearch))
```

**e)** The result of the Friedman test is the same as the ANOVA test: we reject the $H_0$ mentioned before, thus, there is a difference in search times.

```{r}
friedman.test(search$time,search$interface,search$skill)
```

**f)** The one-way ANOVA returns no significant difference in the search time between the interfaces. This result is not very useful, because (1) we removed a lot of information from the model and (2) the model now assumes that the block is a random selection of all available blocks, which is not the case because the blocks were fixed/predetermined.

```{r}
aovsearch2=lm(time~interface, data=search)
anova(aovsearch2)
```

## Exercise 3

**a)** First, we evaluate normality to determine which test to use. From the QQ-plots below, we conclude that both treatment samples are normally distributed. There are three hypotheses: (1) $H_0$ (id) : there is no difference in milk production between cows, (2) $H_0$ (per) : there is no difference in milk production in different periods and (3) $H_0$ (treatment) : there is no difference in milk production with different treatment. From the ANOVA results below, we can conclude that within-cow variation (see variable "id") the milk production differs. Because the p-value for id is less than the significance level of 0.05, therefore, the first $H_0$ is rejected. Furthermore, from the summary, we can conclude that most of the cows (except id4) are different from the cow with id1. Afterwards, we could see p-value for per is less than 0.05, so we reject the second $H_0$ which means that whether a cow is going through the first period or second seems to make a difference. Furthermore, as the p-value of treatment is equal to 0.51654, we do not reject the third $H_0$. This means that treatment A does not significantly differ from treatment B. This could be seen from the second table (treatment B). The summary below indicates that there is no significant difference in milk production. However, it is important to note that this is not the appropriate way of testing the cross-over design.

```{r fig.width=8, fig.height=3}
cow = read.table("cow.txt", header = TRUE);
#There are two factors that influence the milk production
mm = subset(cow, treatment=="A")["milk"]
sf = subset(cow, treatment=="B")["milk"]
mm1 = as.numeric(unlist(mm))
sf1 = as.numeric(unlist(sf))
par(mfrow=c(1,2))
qqnorm(mm1)
qqnorm(sf1)
cow$id = factor(cow$id); cow$per = factor(cow$per)
cowanova = lm(milk~id+per+treatment,data = cow)
anova(cowanova)
summary(cowanova)
```

**b)** In this exercise, we model the cows effect as a random effect by using the function lmer.

```{r}
library(lme4)
cow$order=factor(cow$order)
cowlmer = lmer(milk~treatment+order+per+(1|id), data=cow, REML = FALSE)
summary(cowlmer)
```

Based on the p-value below, we do not reject $H_0$ for treatment. This means that treatment is not important. The result is the same as the result a).

```{r}
cowlmerTreatment = lmer(milk~order+per+(1|id), data=cow, REML = FALSE)
anova(cowlmerTreatment, cowlmer)
```

Based on the p-value below, we do not reject $H_0$ : there is no difference in milk production if the order differs. Therefore, the order of treatment AB is not important.

```{r}
cowlmerOrder = lmer(milk~treatment+per+(1|id), data=cow, REML = FALSE)
anova(cowlmerOrder, cowlmer)
```

Based on the p-value below, we reject $H_0$ for per, which means whether a cow is going through the first treatment or second is important.

```{r}
cowlmerPer = lmer(milk~treatment+order+(1|id), data=cow, REML = FALSE)
anova(cowlmerPer, cowlmer)
```

**c)** From the result below we conclude that the p-value is equal to 0.8281 and we do not reject $H_0$ for treatment. This means that the treatment is not important. From previous analysis, the same conclusion was reached. Given the design, it is inappropriate to use the paired t-test. Since the previous analysis shows that factors such as per have a significant effect on the milk production, it might be unwise to ignore such factors.

```{r}
attach(cow)
t.test(milk[treatment=="A"],milk[treatment=="B"],paired=TRUE)
```

## Exercise 4

**a)** <!-- Make a data.frame in R consisting of two columns and 304 rows. One column should contain an indicator whether or not the patient in that row suffered from nausea, and the other column should indicate the medicin. Make sure these columns match correctly. Study the outcome of xtabs(∼medicin+naus).--> We created a data frame that contains two columns: nausea and medicine. For patients who suffer from nausea, the variable nausea is set to 1. For patients who do not suffer from nausea, the variable is set to 0. We validated that the newly constructed data frame is similar to the dataset by comparing the output of the xtabs function to the original data.

```{r}
data=read.table("nauseatable.txt",header=TRUE)
nausea=c(rep(0,times=100),rep(1,times=52),rep(0,times=32),rep(1,times=35),rep(0,times=48),
  rep(1,times=37))
medicine=c(rep("Chlorpromazine",times=152),rep("Pentobarbital (100mg)",times=67),
  rep("Pentobarbital (150mg)",times=85))
patientdata=data.frame(nausea,medicine)
xtabs(~medicine+nausea,data=patientdata)
```

**b)** <!-- Perform a permutation test in order to test whether the different medicines work equally well against nausea. Permute the medicin labels for this purpose. Use as test statistic the chisquare test statistic for contingency tables, which can be extracted from the output of the command chisq.test: chisq.test(xtabs(∼medicin+nausea))[[1]] --> We perform a permutation test to test $H_0$ :  the medicines work equally well against nausea. To perform this test, we create 1000 random samples (permutations) of the medication data in which the order varies. We match these permutations of the medication data to the ordered nausea data and calculate the chi-square test statistic. The results of the test statistics of the permutations are shown in the histogram below. The result of the chi-square test statistic of the original data is 6.63. We observe that this value deviates from most values in the histogram. To confirm this, we calculated that the p-value is equal to 0.029. The p-value is smaller than the significance level of 0.05, therefore, $H_0$ is rejected and we can conclude that the different medicines do not work equally well against nausea.

```{r fig.width=6, fig.height=3}
medication=factor(patientdata$medicine)
B=1000
tstar=numeric(B)
for (i in 1:B) {
  patientstar=sample(medication) # permute medicine lables
  tstar[i] = chisq.test(xtabs(~patientstar+patientdata$nausea))[[1]] }
hist(tstar)
test_statistic=chisq.test(xtabs(~patientdata$medicine+patientdata$nausea))[[1]]
pl=sum(tstar<test_statistic)/B
pr=sum(tstar>test_statistic)/B
```

**c)** <!-- Compare the p-value found by the permutation test with the p-value found from the chisquare test for contingency tables. Explain the difference/equality of the two p-values. --> When performing the chi-squared test, we observe that the p-value is equal to 0.03643. This is smaller than the significance level of 0.05, therefore, $H_0$ (the medicines work equally well against nausea) is rejected. This is a similar outcome to the permutation test, however, the p-values do differ. The values lie close to each other as both tests are valid in this situation and compare the proportion of expected outcomes (when assuming no dependence) to the actual outcome. The difference can be explained by the different way of calculating the p-value. Using the permutation test, the p-value is calculated by calculating the number of times that the test statistics of the permuted data is smaller than the test statistic of the original data. Using the chi-squared test, the p-value is calculated by comparing the proportion of rows and columns of independend data to the proportion of rows and columns of the actual data.

```{r}
chisq.test(xtabs(~patientdata$medicine+patientdata$nausea))
```

## Exercise 5
In our regression analysis, the response variable is "expend" and the explanatory variables are: "bad, crime, lawyers, employ and pop". The purpose is to explain expend by a numerical function of the explanatory variables.

**a)**
<!-- Make some graphical summaries of the data -->
First, we make a graphical summary of the data by plotting each variable against the others. Furthermore, we construct a histogram of all the numerical variables. Looking at the plots, we observe that expend, lawyers, employ and pop all approximate a linear relationship with each other. Furthermore, state and crime have nonlinear relationhsips with all the other variables. Lastly, the variable bad can be argued to have a weak linear relationship with the variables expend, lawyers, employ and pop. Looking at the histograms, it is interesting to see that almost all variables (expend, bad, lawyers, employ and pop) follow a similar pattern, namely, the lowest value appears frequently and as the value increases, the frequency decreases steeply. Except for a few outliers of frequently occuring high values. In contrast, the variable crime shows a different pattern. Namely, the values in the middle occur also relatively frequently. But the rule: as the value increases, the frequency decreases, applies as well.
<!-- Scatter plots -->
```{r fig.width=8, fig.height=4, echo=FALSE}
data=read.table(file="expensescrime.txt",header=TRUE)
plot(data)
```
<!-- Histograms -->
```{r fig.width=8, fig.height=2, echo=FALSE}
par(mfrow=c(1,ncol(data)-1)) # minus the state column
for (i in 2:ncol(data)) hist(data[,i],main=names(data)[i])
```

To build an intuition of the linear relationship between the depandable variable (expand) and its explanatory variable, we have plotted the explanatory variables against the dependaple variable below. These are the same plots as from the first plot with all factors plotted against each other, but now we can take a closer look. From the plots we can see a strong linear relationship between the dependable expend and 4 of the five factors: bad, lawyers, employ and pop. However we see very big outliers, which will have to be adjusted for, otherwise they will have a very big weight in the factor coefficient from the linear regression.
```{r fig.width=8, fig.height=4, echo=FALSE}

par(mfrow=c(1,3))
plot(data$expend, data$bad)
plot(data$expend, data$crime)
plot(data$expend, data$lawyers)
plot(data$expend, data$employ)
plot(data$expend, data$pop)
```
<!-- Investigate the problem of potential and influence points-->
A potential point is an outlier in an explanatory variable. The effect can be studied by fitting the model with and without the potential point. If the estimated parameters change drastically when removing the potenetial point, the observation is called an influence point. Using the Cook's formula, the distance of an observation on the predictions can be calculated. Whenever the Cook's distance for an observation approximates or is larger than 1, the observation can be considered to be an influence point. As we have not constructed a model yet, we analyse the potential and influence points of our chosen model in c).
<!-- Investigate the problem of collinearity-->
Another relevent concept is collinearity. This is the problem of linear relations between explanatory variables. Collinearity can be detected by a straight line in a scatter plot or by calculating the correlation coefficient. Looking at the scatter plots of the data, we suspect collinearity between the variables expend, lawyers, employ and pop. We confirm this by calculating the correlaction coeffiecients of all possible variable combinations. Looking at the output below, we observe that all the combinations of the variables expend, lawyers, employ and pop have a correlation coefficient above 93. Thus, we conclude that these variables have a collinear relation. The variable bad has a weaker collinear relation with the variables expend, lawyers, employ and pop, namely, ranging from 0.83 to 0.93. Lastly, the variable crime has no collinear relation with any of the other variables. When collinearity is detected among variables, we should avoid having both explanatory variables in the model.
```{r echo=FALSE}
round(cor(data[,2:7]),2)
```
**b)**
<!-- Fit a linear regression model to the data -->
<!-- Use both the step-up and the step-down method to find the best model -->
To fit a linear regression model to the data, first, we start with the step-up method. Using this method, we start by fitting all possible simple linear regression models and calculate the determination coefficient ($R^{2}$). The results are shown in the table below. Looking at this table, we observe that employ has the largerst value of $R^{2}$ (0.954) and is thus selected. Next, we combine this variable with all the variables that do not have a collinear relation with employ. These are the explanatory variables bad and crime. Note that the variable bad still can be considered to be linearly correlated to employ. Adding the variables bad and crime to the models yields in $R^{2}$ = 0.9551. This is an improvement compared to the previous model. Therefore, we continue to add the other vairables to the model. For both models, there is just one variable to add. This results in the last possible option: a model of employ, bad and crime combined. This result in $R^{2}$ = 0.9568, the highest value so far. As there are no more variables to add, the method stops here. The resulting model is: expend = -2.857$e^{+02}$ + 4.979$e^{-02}$ * employ - 1.391$e^{+00}$ * bad + 3.810$e^{-02}$ * crime + error. We have to be carefull as it could be argued that the variables employ and bad are collinear. Therefore, the model that is constructed with the variables employ and crime (expend = -2.484$e^{+02}$ + 4.630$e^{-02}$ * employ + 2.962$e^{-02}$ * crime + error) might be a better model as it contains fewer variables and the value of $R^{2}$ is similar.

| **Explantory Variable(s)** | bad    | crime  | lawyers | employ | pop    | bad, employ | crime, employ | bad, crime, employ |
|---------------------|--------|--------|---------|--------|--------| -------------|--------------- | --------------- |
| **Multiple R-squared**  | 0.6964 | 0.1119 | 0.9373  | 0.954  | 0.9073 | 0.9551       | 0.9551         | 0.9568  |

```{r echo=FALSE,results='hide'}
summary(lm(expend~bad,data=data))
summary(lm(expend~crime,data=data))
summary(lm(expend~lawyers,data=data))
summary(lm(expend~employ,data=data))
summary(lm(expend~pop,data=data))
summary(lm(expend~bad+employ,data=data))
summary(lm(expend~crime+employ,data=data))
summary(lm(expend~bad+crime+employ,data=data))
```
Second, we use the step-down method. This method start with fitting all explanatory variables in the so-called full model. In each iteration, one explanatory variable is removed. This time, we try the model with all variables, regardless of collinearity. In round 1, we observe that the variable crime has the highest p-value, 0.25534 > 0.05, therefore, the variable crime will be removed. In round 2, pop has the highest p-value, 0.06012 > 0.05, therefore, the variable pop will be removed. In round 3, bad has the highest p-value, 0.34496 > 0.05, therefore, the variable bad will be removed. In round 4, lawyers has the highest p-value, 0.00113 < 0.05, therefore, the variable will not be removed and the method stops. This results in the model expend = -1.107$e^{+02}$ + 2.686$e^{-02}$ * lawyers + 2.971$e^{-02}$ * employ + error.

**Round 1: expend ~ bad, crime, lawyers, employ, pop**

| **Explantory Variables** | bad | crime | lawyers | employ | pop | 
|--------------------------|-----|-------|---------|--------| ----|
| **p-value**              |0.02719|0.25534|0.00592|0.00354|0.03184|

**Round 2: expend ~ bad, lawyers, employ, pop**

| **Explantory Variables** | bad | lawyers | employ | pop | 
|--------------------------|-----|---------|--------| ----|
| **p-value**              |0.05402|0.00106|0.00380|0.06012|

**Round 3: expend ~ bad, lawyers, employ**

| **Explantory Variables** | bad | lawyers | employ |
|--------------------------|-----|---------|--------|
| **p-value**              |0.34496|0.00147|1.2$e^{-06}$|

**Round 4: expend ~ lawyers, employ**

| **Explantory Variables** | lawyers | employ |
|--------------------------|---------|--------|
| **p-value**              |0.00113|4.89$e^{-07}$|

```{r echo=FALSE,results='hide'}
summary(lm(expend~bad+crime+lawyers+employ+pop,data=data))
summary(lm(expend~bad+lawyers+employ+pop,data=data))
summary(lm(expend~bad+lawyers+employ,data=data))
summary(lm(expend~lawyers+employ, data=data))
```
<!-- If step-up and step-down yield two different models, choose one and motivate your choice -->
Using the step-up and step-down method resulted in two different models. The advantage of the model constructed using the step-up method (expend ~ employ bad crime) is that the variables are not collinear. The advantages of the model constructed using the step-down method (expend ~ lawyers employ) are that the value of $R^{2}$ (0.9632) is higher compared to the $R^{2}$ value of the step-up model (0.9568) and the model contains fewer explanatory variables. However, as the collinearity of variables weight higher compared to number of variables and the difference of $R^{2}$ is relatively small, we prefer the model constructed using the step-up model. We even consider removing the variable bad of the step-up model as it can be argued to have a collinear relation with the variable employ.

**c)**
<!-- Check the model assumptions by using relevant diagnostic tools -->
We check the model (expend ~ employ crime) assumptions (linearity of the relation and normality of the errors) using both graphical and numerical tools.
<!-- #1 Scatter plot: plot Y against each X k separately -->
First, we look at the scatter plot of the response variable against each explanatory variable separately. This is visualized in a). For each combination with expend, we see two outliers at the right of the scatter plots. Moreover, the relation with the variables bad, lawyers, employ and pop is linear and nonlinear combined with state and crime. 
<!-- #2 Scatter plot of residuals against each Xk in the model separately -->
Second, we construct the scatter plot of the residuals against each explanatory variable that is in the model (crime and employ) seperately. The plot with crime contains a cluster around the middle line and the plot with employ has a cluster in the bottom left diagnal with two outliers in the upper right corner. This means...
<!-- #3 Added variable plot of residuals of Xj against residuals of Y with omitted Xj -->
Third, we construct the added variable plot. In this plot, the residuals of the explanatory variables are plotted against the residuals of the model without that specific variable. This shows the effect of adding an explanatory variable to the model. Looking at the figure, we observe that the plots for bad, lawyers and pop contain compact clusters. The plots of crime and employ are more spread. This means...
<!-- #4 Scatter plot of residuals against each Xk not in the model separately -->
Next, we construct the scatter plots of the residuals against each explanatory variable that is not in the model (bad, lawyers and pop) seperatly. The outputs are similar to each other. When looking at the pattern we do not observe a linear relation and therefore should not include more variables.
<!-- #5 Scatterplot of residuals against Y and Y^ -->
Afterwards, we construct the scatter plot of the reduals against the response variable and the fitted model. We observe little spread as in both plots there is a cluster around zero with only few outliers. This means...
<!-- #6 normal QQ-plot of the residuals and Shapiro-Wilk's method -->
Lastly, we check the normality assumption by constructing a qq-plot and conduction Shapiro-Wilk's test. We cannot assume normality as the qq-plot does not approximate a straight line, this means that the model is invalid. Unfortunetely, none of the other models resulted in a normal distribution of the residuals. 

```{r fig.width=8, fig.height=4, echo=FALSE}
expendlm = lm(expend~crime+employ,data=data)
par(mfrow=c(2,4))
#2 Scatter plot of residuals against each Xk in the model separately.
plot(residuals(expendlm),data[,4],xlab="residuals",ylab="crime")
plot(residuals(expendlm),data[,6],xlab="residuals",ylab="employ")
#4 Scatter plot of residuals against each Xk not in the model separately
plot(residuals(expendlm),data[,3],xlab="residuals",ylab="bad")
plot(residuals(expendlm),data[,5],xlab="residuals",ylab="lawyers")
plot(residuals(expendlm),data[,7],xlab="residuals",ylab="pop")
#5 Scatterplot of residuals against Y and Y^
plot(residuals(expendlm),data[,2],xlab="residuals",ylab="expend")
plot(residuals(expendlm),fitted(expendlm),xlab="residuals",ylab="fitted model")

par(mfrow=c(2,3))
#3 Added variable plot of residuals of Xj against residuals of Y with omitted Xj
x = residuals(lm(bad~employ+crime,data=data))
y = residuals(lm(expend~employ+crime,data=data))
plot(x,y,main="Added variable plot for bad")
x = residuals(lm(crime~bad+lawyers+pop,data=data))
y = residuals(lm(expend~bad+lawyers+pop,data=data))
plot(x,y,main="Added variable plot for crime")
x = residuals(lm(lawyers~employ+crime,data=data))
y = residuals(lm(expend~employ+crime,data=data))
plot(x,y,main="Added variable plot for lawyers")
x = residuals(lm(employ~bad+lawyers+pop,data=data))
y = residuals(lm(expend~bad+lawyers+pop,data=data))
plot(x,y,main="Added variable plot for employ")
x = residuals(lm(pop~employ+crime,data=data))
y = residuals(lm(expend~employ+crime,data=data))
plot(x,y,main="Added variable plot for pop")

#6 normal QQ-plot of the residuals and Shapiro-Wilk's method
qqnorm(residuals(expendlm))
#shapiro.test(residuals(expendlm))
```

## Appendix: R code
```{r echo=TRUE, eval=FALSE}

# --- Exercise 4 --- #

# A
data=read.table("nauseatable.txt",header=TRUE)
nausea=c(rep(0,times=100),rep(1,times=52),rep(0,times=32),rep(1,times=35),rep(0,times=48),
  rep(1,times=37))
medicine=c(rep("Chlorpromazine",times=152),rep("Pentobarbital (100mg)",times=67),
  rep("Pentobarbital (150mg)",times=85))
patientdata=data.frame(nausea,medicine)
xtabs(~medicine+nausea,data=patientdata)
#B
medication=factor(patientdata$medicine)
B=1000
tstar=numeric(B)
for (i in 1:B) {
  patientstar=sample(medication) # permute medicine lables
  tstar[i] = chisq.test(xtabs(~patientstar+patientdata$nausea))[[1]] }
hist(tstar)
test_statistic=chisq.test(xtabs(~patientdata$medicine+patientdata$nausea))[[1]]
pl=sum(tstar<test_statistic)/B;pl
pr=sum(tstar>test_statistic)/B;pr
#C
chisq.test(xtabs(~patientdata$medicine+patientdata$nausea))

```
