---
title: "Assignment 3 Exercise 3"
author: "Geoffrey"
date: "3/15/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 3

**a)** First, we check if there are any linear correlated factors in the model by creating a scatterplot of all the variables. Looking at the scatterplot below, we conclude that there are no linear correlations. Afterwards, using the generalised linear regression model function, we run the Poisson regression. The output is presented below.

```{r fig.width=8, fig.height=5, echo=FALSE}
africa = read.table("africa.txt", header = TRUE)
plot(africa)
```

```{r echo=FALSE}
africa$pollib = factor(africa$pollib)
africaglm=glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numelec+numregim, family=poisson,data=africa)
summary(africaglm)
```

We conclude that oligarchy, pollib and parties significantly estimate (or have a linear relation with) the amount of successful military coups. As we take pollib as a factor, we find that category 2 (full civil rights) has significant less military coups (estimated 1.69 coups less) than pollib category 0. Afterwards, to evaluate the model, we plotted the residuals against the fitted values. The plot shows equal variance, however, a pattern can be observed. This is due to the dependent variable being a count on a small scale (0 - 6) which can be interpreted as discrete data. Approximately, for each target value, a curve is visible. Next, we calculate the logarithm to ensure that the x-values are fitted by a linear function. The second plot shows more spread, however, the previously mentioned structure of curves is still visible. Finally, we plot the response residuals. We observe that the response residuals increase with the (logarithm) of the fitted values, as expected under a Poisson model. 

```{r fig.width=8, fig.height=3, echo=FALSE}
par(mfrow=c(1,3))
plot(fitted(africaglm),residuals(africaglm))
plot(log(fitted(africaglm)),residuals(africaglm))
plot(log(fitted(africaglm)),residuals(africaglm, type="response"))
```

**b)** Following the step down method, we removed the factors in the order: numelec > numregim > size > popn > pctvote. This results in the model miltcoup = 0.251377 + 0.092622 * oligarchy - 0.574103 * pollib + 0.022059 * parties + error. In this process, we started with an R-squared value of 0.57 and ended up with a value of 0.50, however, we reduced the model from eight factors to three. Moreover, the residual plots look similar to the ones in a) in which all factors were included in the model.

```{r echo=FALSE, results='hide'}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numelec+numregim, family=poisson,data=africa))
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numregim, family=poisson,data=africa))
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size, family=poisson,data=africa))
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn, family=poisson,data=africa))
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote, family=poisson,data=africa))
summary(glm(miltcoup~oligarchy+pollib+parties, family=poisson,data=africa))
```
```{r fig.width=8, fig.height=3, echo=FALSE, results='hide'}
africaglm2=glm(miltcoup~oligarchy+pollib+parties, family=poisson,data=africa)
with(summary(africaglm2), 1 - deviance/null.deviance)
summary(africaglm2)
par(mfrow=c(1,3))
plot(fitted(africaglm2),residuals(africaglm2))
plot(log(fitted(africaglm2)),residuals(africaglm2))
plot(log(fitted(africaglm2)),residuals(africaglm2, type="response"))
```