---
title: "Assignment 3 Exercise 3"
author: "Geoffrey"
date: "3/15/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 3

**a)**
First we check if there are any linear correlated factors in the model:
```{r}
africa = read.table("africa.txt", header = TRUE)
plot(africa)
```

We conclude that there are no linear correlations.

With the generalised linear regression model function  we run the poisson regression with the following result:
```{r}
africaglm=glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numelec+numregim, family=poisson,data=africa)
summary(africaglm)
```

We conclude that oligarchy, pollib and parties signifcantly estimate (or have a linear relation with) the amount of successful military coups. Furthermore the plotted residuals against the fitted values show kind of equal variances. However a pattern can be seen, but I am not sure what this indicates (as we are not familiar enough with deviance residuals).

```{r}
plot(log(fitted(africaglm)),residuals(africaglm))
```
 

**b)**
In the step down method we have removed the following factors in the order: fanumelec > numregim > size > popn > pctvote. With this process we went from a R squared value of 0.5652689 to 0.5017707, but reduced the formula from eight factors to three. The fitted values against the residuals look the same as before.
```{r}
africaglm2=glm(miltcoup~oligarchy+pollib+parties, family=poisson,data=africa)
with(summary(africaglm2), 1 - deviance/null.deviance)
summary(africaglm2)
plot(log(fitted(africaglm2)),residuals(africaglm2))
```