---
title: "A3_EX1"
author: "yizhen"
date: "3/13/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=8, fig.height=4) 
```

## Exercise 1

**a)** 
<!-- Make an informative plot of the data. --> First, we add a column 'loglongevity' which will be used as a response variable. Next, we plot the longevity data in a separate boxplot for each activity. We observe that the longevity for fruitflies of the activity 'isolated' is the longest, followed by the activity 'low' and the activity 'high' has the lowest longevity. Looking at the scater plot of loglongevity and thorax, we observe a weak linear correlation. The points follow a linear pattern, however, they are relatively widely spread. Third, we could observe a weak linear correlation bewteen longevity and thorax.

```{r echo = FALSE, results='hide'}
fliesdata = read.table("fruitflies.txt", header = TRUE)
fliesdata$loglongevity = log(fliesdata$longevity)
attach(fliesdata)
par(mfrow=c(1,2))
plot(longevity~activity, data=fliesdata)
plot(loglongevity~thorax, data=fliesdata);abline(lm(loglongevity~thorax, data=fliesdata))
pairs(fliesdata)
fliesanova = lm(loglongevity~activity, data=fliesdata)
anova(fliesanova)
```

<!--Investigate whether sexual activity influences longevity by performing a statistical test, without taking the thorax length into account.--> 
In order to investigate whether sexual activity influences longevity, we performed an one-way Anova test. The null hypothesis states that sexual activity does not influence the longevity. The test results in a p-value smaller than the significance level of 0.05. Therefore, we reject $H_0$ and thus conclude that the sexual activity will influence the longevity. 
<!--What are the estimated longevities for the three conditions? Comment.-->
According to the summary, the estimated longevity for group 'high' is 3.60, and for group 'isolated' 3.60 + 0.52 = 4.12 and for group 'low' 3.60 + 0.39 = 3.99. With a 95% confidence interval, the logevity for 'high' is [3.48 3.72], for 'isolated' [3.82, 4.41] and for 'low' [3.70, 4.29]. From this, we confirm that a high sexual acitivty has a negative impact on the longevity.
```{r echo = FALSE}
summary(fliesanova)
confint(fliesanova)
```

**b)** 
<!--Investigate whether sexual activity influences longevity by performing a statistical test, now including thorax length as an explanatory variable into the analysis. --> For this exercise, we apply two-way Anova with the two factors: activity and thorax. With $H_0$ (1) activity does not influence longevity, (2) thorax does not influence logevity, and (3) there is no interaction between activity and thorax. The output of this test shows that the p-values for the first two null hypotheses are all smaller than 0.05, therefore, we reject the first two null hypotheses. This means that activity and thorax influence the longevity. The p-value for the third null hypothesis is 0.4574. This is larger than 0.05, therefore, we do not reject the third null hypothesis. This means that there is no interaction betweem them. Then our model fit the additive model.

```{r echo = FALSE, results='hide'}
fliesdata$activity = as.factor(fliesdata$activity)
fliesdata$thorax = as.numeric(fliesdata$thorax)
fliesanova = lm(loglongevity~thorax*activity, data=fliesdata)
anova(fliesanova)
```

<!-- Does sexual activity increase or decrease longevity?  What are the estimated longevities for the three groups, for a fly with average thorax length? -->
Now from the result we could know the p-values for both activity and thorax are smaller than significant level 0.05. Therefore $H_0$ here are rejected which means acitivity and thorax will effect the longevity. 
<!--Need review !! i changed thorax to fliesdata$thorax = as.numeric(fliesdata$thorax), is it correct now?  -->
We calculated the mean of thorax equal to 0.82 and from summary we could see the estimated throax is 0.067. And activity3 = -(activity1+activity2) = 0.054. For 'high': 0.82+3.309-0.233+0.067=3.963, for 'isolated': 0.82+3.309+0.179+0.067=4.375, for 'low': 0.82+3.309+0.054+0.067=4.25.
According to the result, we conclude that the higher activity is, the shorter longevity they have, the result is similar in a). 

```{r echo = FALSE, results='hide'}
mean(thorax,data=fliesdata)
fliesdata$activity = as.factor(fliesdata$activity)
fliesdata$thorax = as.numeric(fliesdata$thorax)
fliesanova2 = lm(loglongevity~thorax+activity, data=fliesdata)
anova(fliesanova2)
```

```{r echo = FALSE}
contrasts(fliesdata$activity)=contr.sum;
fliesanova2 = lm(loglongevity~activity+thorax, data=fliesdata)
summary(fliesanova2)
```

**c)**
<!-- How does thorax length influence longevity? Investigate graphically and by using an appropriate test whether this dependence is similar under all three conditions of sexual activity. -->
From the graph below we could see that longevity increase with the throax. Group 'isolated' has the longest longevity, followed by 'low' and 'high'.

```{r echo = FALSE}
plot(fliesdata$loglongevity~fliesdata$thorax,pch=as.character(fliesdata$activity))
```

Because thorax will influence the longevity, its dependence on activity is not so clear. Here we apply ANCOVA. Using 'drop1' to get the p-value. According to result, p-values are all less than significance level 0.05. It confirms our analysis before that both activity and thorax will influence the longevity.

```{r echo = FALSE, results='hide'}
drop1(fliesanova2, test="F")
contrasts(fliesdata$activity)=contr.sum
fliesanova2 = lm(loglongevity~thorax+activity, data=fliesdata)
summary(fliesanova2)
```

From the plot and summary below we could see that p-values for 'isolated:thorax' and 'low:thorax' are bigger than significance level 0.05, therefore we do not reject $H_0$ here which is there is no difference on thorax's dependence under three activities. So the dependence is similar under all three conditions of sexual activity.

```{r echo = FALSE}
plot(loglongevity~thorax, pch=unclass(activity));abline(lm(loglongevity~thorax, data=fliesdata[fliesdata$activity == 'isolated',]), col='blue'); abline(lm(loglongevity~thorax, data=fliesdata[fliesdata$activity == 'high',]), col='green'); abline(lm(loglongevity~thorax, data=fliesdata[fliesdata$activity == 'low',]), col='red')
```

```{r echo = FALSE}
aovfliesinter = lm(loglongevity~activity*thorax, data=fliesdata)
summary(aovfliesinter)
```

**d)** 
<!-- Which of the two analyses, without or with thorax length, do you prefer? Is one of the analyses wrong?  -->
We prefer to take thorax length into account, due to our analysis above, we know that thorax will influence the longevity of fruitflies. So it is not wise to ignore such a factor when doing analysis. But the first analysis is not wrong. At the begining, we don't know thorax's effect towards longevity and we only take one factor(activity) into account. Therefore, we apply one-way anova. They all get us right results. As the first one only focus on activities' influence to longevity and second one focus on both activity and thorax. 

**e)**
<!-- Verify normality and heteroscedasticity by making a normal QQ-plot of the residuals, and a residuals versus fitted plot, for the analysis that includes thorax length. -->
In QQ plot we conclude that normality is ok. For the residuals versus fitted plot there is no clear pattern therefore we conclude that there is no sign of heteroscedasticity.

```{r echo = FALSE}
par(mfrow=c(1,2))
qqnorm(residuals(fliesanova2));qqline(residuals(fliesanova2))
plot(fitted(fliesanova2), residuals(fliesanova2))
```

**f)**
<!-- Perform the ancova analysis with the number of days as the response, rather than its logarithm. Verify normality and heteroscedasticity of the residuals of this analysis. Was it wise to use the logarithm as response? -->
We do the same ancova analysis but use longevity as response variable. From the result we could know p-values for thorax and activity are smaller than significance level 0.05 therefore we get same conclusion as before that thorax and activity will effect fruitflies' longevity. Also we could see from the first plot that longevity increase with thorax. Then from the qq plot we could see the normality is also good. And from residuals versus fitted plot, we noticed some pattern and residuals seem to be bigger with bigger fitted values. So the inference here is, heteroscedasticity exists. 
In conclusion, it is wise to use the logarithm as response as we don't see heteroscedasticity in that model. 

```{r echo = FALSE, results='hide'}
fliesanova3 = lm(longevity~thorax+activity, data=fliesdata)
drop1(fliesanova3, test="F")
contrasts(fliesdata$activity)=contr.sum
fliesanova3 = lm(longevity~thorax+activity, data=fliesdata)
summary(fliesanova3)
```

```{r echo = FALSE}
plot(fliesdata$longevity~fliesdata$thorax, pch=as.character(fliesdata$activity))
par(mfrow=c(1,2))
qqnorm(residuals(fliesanova3));qqline(residuals(fliesanova3))
plot(fitted(fliesanova3), residuals(fliesanova3))
```
